---
title: "Activity_Monitoring"
author: "Yukun Liu"
date: "3 September 2016"
output: html_document
---

## Data file preparation
Load the required libraries and data file.
```{r echo=TRUE}
# Load libraries
library(dplyr)
library(ggplot2)
# Output the figures to folder figure.
knitr::opts_chunk$set(fig.path = "figure/")
# Set working directory and prepare data file. If the file doesn't exist, start downloading.
setwd("~/Documents/Data Scientiest_Course/Reproducible")
destfile <- "activity_monitoring_data.zip"
if (!file.exists(destfile)) {  
        # if destfile doesn't exist, start downloading the file from the provided url.
        fileurl <- "https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2Factivity.zip"
        download.file(fileurl, destfile, method = "curl")
        unzip(destfile)
}
```

## Loading and preprocessing the data

Using read.csv() to read the csv file activity.csv into raw data frame raw_activity. 
Data frame activity is used to exclude missing values (steps = NA).

```{r echo=TRUE}
raw_activity <- read.csv("activity.csv", na.strings = "NA")

# Remove rows with steps is na.
activity <- raw_activity[!is.na(raw_activity$steps),]
# Show the preprocessed data
summary(activity)
head(activity)
dim(activity)
```

## What is the total number of steps taken per day?
Using dplyr package, we can group data frame activity and calculate the steps per day. The result is saved into data frame steps_per_day

```{r echo=TRUE}

steps_per_day <- activity %>% 
        group_by(date) %>%
        summarise(steps = sum(steps)) %>%
        data.frame()
# Histogram for steps per day
hist(steps_per_day$steps, breaks = 10, ylim = c(0, 25), 
     main="Total Number of Steps Taken Per Day")

# mean of the total number of steps per day
mean(steps_per_day$steps)
# mean of the total number of steps per day
median(steps_per_day$steps)
```

## What is the average daily activity pattern?

```{r echo=TRUE}
# Calculate the average steps per interval across all days.
steps_per_interval <- activity %>%
        group_by(interval) %>%
        summarise(steps = mean(steps)) %>%
        data.frame()

# Convert the interval variable into time format
x <- strptime(formatC(steps_per_interval$interval, width=4, flag="0"), "%H%M")
y <- steps_per_interval$steps

# Plot the line chart
plot(x, y, type = "l", main = "Average Daily Activity Pattern", 
     xlab="Time", ylab="Average Steps", col="red")

# Locate the interval with maximum number of steps
df <- data.frame(interval = x, steps = y)
xmax <- strftime(df[df$steps==max(df$steps),]$interval, "%H:%M")
xmax

```

## Imputing missing values
Note that there are a number of days/intervals where there are missing values (coded as ð™½ð™°). The presence of missing days may introduce bias into some calculations or summaries of the data.

```{r echo=TRUE}
# Calculate the total number of missing values in the raw data set raw_activity

missing_steps <- raw_activity[is.na(raw_activity$steps),]
length(missing_steps$steps)
# Imputing the missing value with mean steps of the interval where the missing value is located

avg_steps <- data.frame(steps_per_interval)

missing_steps.avg <- merge(missing_steps, avg_steps, by = "interval")
steps <- missing_steps.avg$steps.y
date <- missing_steps.avg$date
interval <- missing_steps.avg$interval

# Data frame with all missing values filled in with average steps per interval
activity.completed <- rbind(activity, data.frame(steps, date, interval))

# Re-calculate total number of steps taken per day 
steps_per_day <- activity.completed %>% 
        group_by(date) %>%
        summarise(steps = sum(steps)) %>%
        data.frame()

# Histogram for steps per day with missing values filled
hist(steps_per_day$steps, breaks = 10, ylim = c(0, 25), 
     main="Total Number of Steps Taken Per Day (Imputing Missing Values)")
# mean of the total number of steps per day
mean(steps_per_day$steps)
# mean of the total number of steps per day
median(steps_per_day$steps)
# 
```
The missing values had slight impact on the median steps per day and no impact on the average (mean) steps per day if filled with average (mean) steps per interval.

However, the total number of steps taken per day would be notably increased with imputing missing values according to the revised histograms. 

## Pattern differences between weekdays and weekends

```{r echo=TRUE}

wdays <- c("Mon", "Tue", "Wed", "Thu", "Fri")

activity.completed$day <- weekdays(as.Date(activity.completed$date), abbreviate = TRUE)

activity.completed$wday <- factor((activity.completed$day %in% wdays), 
                                  levels=c(FALSE, TRUE), labels=c('Weekend', 'Weekday')
                                  )
# Calculate the average steps per interval across all days.
steps_per_interval <- activity.completed %>%
        group_by(interval, wday) %>%
        summarise(steps = mean(steps)) %>%
        data.frame()

# Convert the interval variable into time format

x <- steps_per_interval$interval
y <- steps_per_interval$steps
wday <- steps_per_interval$wday

df <- data.frame(interval = x, steps = y, wday = wday)
# Plot the line chart with ggplot

p <- ggplot(df, aes(x = interval, y = steps))
p + geom_line(color="blue") + facet_grid(wday ~ .) + 
        ggtitle("Average Daily Activity Pattern") + 
        xlab("Interval") + ylab("Number of Steps") 
```